\documentclass[10pt]{article}

\input{defs}

\title{\bfseries{The AppleCore Language Specification, v1.0}}
%
\author{Robert L. Bocchino Jr.\\
Pittsburgh, PA}

\begin{document}

\maketitle

\tableofcontents

\section{Introduction and Rationale}

AppleCore is a C-like programming language for the Apple II series of
computers.  A cross-compiler from Java (e.g., running on a Mac or PC)
to the Apple II is under development.

The idea of AppleCore is to provide a ``core language for programming
the Apple II.''  The motivation is as follows:
%
\begin{enumerate}
%
\item Programming the Apple II is both nostalgic and just plain fun.
%
\item The Apple II is a small machine.  Therefore it needs a
  lightweight programming model (i.e., without complicated runtime
  overhead in space or time).  The programming model must also provide
  (1) tight control over the layout of generated code and (2) easy
  integration with assembly-language support routines such as those
  provided by the System Monitor or Applesoft.
%
\item The native programming models on the Apple II are inadequate.
  Assembly language is powerful but painfully low-level.  Applesoft,
  on the other hand, is a disaster.  It lacks even the most basic
  abstraction required of a sane programming environment, including
  scoped loops, scoped functions with local variables, and variable
  names longer than \emph{two letters}!  Applesoft does have nice
  support for numeric and string input and output, so if you just want
  to write short programs that read and write things, it's handy. But
  for any but the tiniest programs, Applesoft code becomes an
  unreadable mess.  In terms of control abstraction Applesoft is much
  worse than assembly language, because it forces you to branch to
  numbered lines, whereas in assembly language you can \kwd{JMP} and
  \kwd{JSR} to symbolic labels with meaningful names.
%
\end{enumerate}

The goal of AppleCore is to rectify this situation by providing a
``low-level high-level'' language (at about the same level of
abstraction as C, but without structs) that's both useful and fun to
program with.  First, it's a compiled language, so it should run
faster than Applesoft.  Second, it's modular and supports separate
compilation. The entire core language needs only a few kilobytes of
support routines; additional routines can be linked in as needed.
Third, AppleCore strikes a balance between power and efficiency.
Notably, it supports integer arithmetic on fixed-size variables of up
to 255 bytes.  This power does come at a cost in terms of speed for
single-byte arithmetic that could be done in registers; however, if
that speed is needed, one can easily call into assembly language
arithmetic routines.  Fourth, it provides easy integration with
assembly language routines: you can say things like ``load `A' into
the accumulator and call \$FDED'' that are impossible to say directly
in Applesoft.

The main things that AppleCore lacks, and that Applesoft provides, are
automatic memory management and floating-point computation.  I've left
out memory management to keep things very simple.  Basically, as a
programmer I can't live happily without control abstractions (loops
and functions); but once I have them I'm perfectly content to
synthesize data abstractions on my own.  If this becomes too much of a
pain, a library for memory management can be added without difficulty,
after the fashion of C.  (Applesoft actually provides garbage
collection, but it works very poorly, causing the machine to seize up
when the memory limit is reached.)  I've left out floating-point
computation mainly because I'm most interested in text processing and
integer computations; but again, FP computations can be added without
too much difficulty.  In the mean time, one can fake FP computations
by encoding FP literals as string data and calling into the Applesoft
FP routines to convert from strings to FP numbers and back, and to
compute using the FP numbers.  Applesoft is actually quite good at
this.

\section{Lexical Structure}

Before parsing a source file, the source text is separated into
\emph{tokens}.  There are four classes of tokens: identifiers,
keywords, constants, and symbols.

\subsection{White Space}
\label{sec:lexical:white-space}

White space consists of any sequence of the following characters:
space (ASCII SP, value \$20), newline (ASCII NL, value \$0A), carriage
return (ASCII CR, value \$0D), and horizontal tab (ASCII HT, value
\$09).  Whitespace is syntactically relevant only in the following
ways:
%
\begin{enumerate}
%
\item \ithead{Separation of tokens.} When a whitespace character or end of
file appears immediately after a non-whitespace character, that
signifies the end of a token.
%
\item \ithead{End of line.} An end-of-line sequence (\kwd{EOL}) is a
  comment terminator (see Section~\ref{sec:lexical:comments}).  As
  usual, the definition of \kwd{EOL} is platform-dependent: on the
  Apple II it is CR, on UNIX it is NL, and on Windows it is CR
  followed by NL.
%
\end{enumerate}
%
The lexer may also count source lines (using \kwd{EOL}) to provide
line numbers for error messages.

\subsection{Identifiers}\label{Identifiers}

An identifier is a sequence of letters, underscore character, and
digits that is not a keyword.  The first character must be a letter.
Upper- and lowercase letters are distinct.  A single identifier may
contain at most 32 characters.

\subsection{Keywords}

The following sequences of characters are reserved for use as keywords
and may not be used as identifiers:

\begin{ttfamily}
\begin{center}
\begin{tabular}{l l l l}
AND & CONST & DATA & DECR \\
%
ELSE & FN & IF & INCLUDE \\
%
INCR & NOT & OR & RETURN \\
%
SET & VAR & XOR & WHILE \\
\end{tabular}
\end{center}
\end{ttfamily}

Following Apple II tradition (mostly because the original Apple II had
no support for lower-case letters), AppleCore keywords are uppercase.
That means that the same words in lowercase (or a combination of
upper- and lowercase) are not recognized as keywords, so those
sequences of characters are available for use as identifiers.

\subsection{Constants}
\label{sec:lexical:constants}

There are three types of constants: integer constants, string
constants, and character constants.  

\bfhead{Integer constants} Integer constants may be written in decimal
or hexadecimal form.  A decimal integer constant consists of one or
more decimal digits \kwd{0} through \kwd{9}.  For example, \kwd{1} and
\kwd{123} are valid integer constants in decimal form.  A hexadecimal
integer constant consists of a dollar sign \kwd{\$} followed by by one
or more hexadecimal digits \kwd{0} through \kwd{9} or \kwd{A} through
\kwd{F}.  For example, \kwd{\$1}, \kwd{\$A}, and \kwd{\$00FF} are
valid integer constants in hexadecimal form.

Integer constants represent values with up to 255 bytes of precision
(i.e., in the range $0$ through $2^{255 \cdot 8}-1$, inclusive).  It
is a compile-time error to write an integer constant with a value
larger than that.

\bfhead{String constants} A string constant is a sequence of
characters enclosed in double quotes (\kwd{"..."}).  A string constant
represents a sequence of ASCII characters, one for each character
appearing in the double-quotes, except that the character sequence
\kwd{\bs\$} has the special meaning described below.

Since not all ASCII characters have printable representations, an
\emph{escape sequence} may be used to represent an arbitrary ASCII
value (printable or non-printable).  An escape sequence consists of a
backslash \kwd{\bs} followed by a dollar sign \kwd{\$} and two
hexadecimal digits.  The whole sequence represents the single
character with the ASCII value given by the digits.  For example, the
string constant

$$\kwd{"Hello, world!\bs\$0D"}$$

\noindent
represents a string consisting of the characters \kwd{Hello, world!}
followed by a CR character.  Similarly, the quote character \kwd{"}
can be embedded in a string constant with the sequence \kwd{\bs\$22}.

\bfhead{Character constants} A character constant consists of a
single-quote character \kwd{'}, followed by a printable ASCII
character, followed by another single-quote character.  It represents
the ASCII value associated with the character.  For example, the
constant \kwd{'A'} represents the value $\$41$.

\subsection{Symbols}

AppleCore uses the symbols shown in Figure~\ref{fig:symbols}, each of
which is a separate token.

\begin{figure}[th]
\begin{center}
\begin{tabular}{c l}
\textbf{Symbol} & \textbf{Meaning} \\
%
\kwd{@} & Denotes the address of a variable, or an address type \\
%
\kwd{\textasciicircum} & Denotes a 6502 register expression \\
\kwd{*}  & Multiplication \\
\kwd{/}  & Division \\
\kwd{+}  & Addition \\
\kwd{-}  & Negation (as unary operator); subtraction (as binary
operator) \\
\shl & Left shift \\
\shr & Right shift \\
\kwd{>=} & Greater than or equal to \\
\kwd{<=} & Less than or equal to \\
\kwd{>}  & Greater than \\
\kwd{<}  & Less than \\
\kwd{=}  & Equal to; assignment \\
\kwd{(} and \kwd{)} & Encloses function parameters and arguments, parenthesized expressions
\\
\kwd{\{} and \kwd{\}} & Encloses statement blocks \\
\kwd{[} and \kwd{]} & Address dereference \\
\kwd{;} & Terminates declarations and statements \\
\kwd{:} & Separates variable declaration from size \\
\kwd{,} & Separates function parameters and arguments; separates index from size in dereference \\
\kwd{\bs} & Signifies unterminated string data \\
\end{tabular}
\end{center}
\caption{Symbols used in AppleCore syntax.}
\label{fig:symbols}
\end{figure}

\subsection{Comments}
\label{sec:lexical:comments}

The character \kwd{\#} indicates a comment; all text to the next end
of line (or end of file, if there is no end of line) are ignored by
the lexer.  Multi-line comments are indicated by preceding each line
with \kwd{\#}.

\section{Syntax}
\label{sec:syntax}

The syntax description below uses the following conventions:
%
\begin{itemize}
%
\item The symbol $^*$ denotes zero or more instances of the entity
  preceding it.
%
\item Italicized parentheses \group{} group the enclosed symbols and
  do not denote program text.
%
\item Italicized brackets \opt{} signify that the enclosed symbols are
  optional (i.e., they may occur zero or one time).  They do not
  denote program text.
%
\item The nonterminal \nonterm{identifier} stands for any identifier
  as defined in Section \ref{Identifiers}.
%
\item The nonterminals \nonterm{integer-const},
  \nonterm{string-const}, and \nonterm{char-const} stand for integer,
  string, and character constants as defined in
  Section~\ref{sec:lexical:constants}.
%
\item The nonterminal \nonterm{size} refers to an integer constant
  whose value is between 1 and 255 (inclusive).
%
\item Text and symbols in \kwd{typewriter} font (including
  non-italicized parentheses and brackets) denote literal program
  text.
%
\end{itemize}

\subsection{Source Files}
\label{sec:syntax:source-files}

The basic syntactic unit of an AppleCore program is a \emph{source
  file}, i.e., an input file presented to the AppleCore compiler for
compilation.  The compiler translates the source file into an assembly
file which is then linked with other assembly files as discussed in
Section~\ref{sec:code-gen:source-files} to form a complete executable
program.

An AppleCore source file is given by zero or more
declarations (Section~\ref{sec:syntax:decls}):

$$\nonterm{source-file} ::= \nonterm{decl}^*$$

\subsection{Declarations}
\label{sec:syntax:decls}

A declaration is a constant declaration, a data declaration, a
variable declaration, a function declaration, or an include
declaration:

$$\nonterm{decl} ::= \mbox{\nonterm{const-decl} $|$
  \nonterm{data-decl} $|$ \nonterm{var-decl} $|$ \nonterm{fn-decl}
    $|$ \nonterm{include-decl}}$$

\bfhead{Constant declarations} A constant declaration consists of the
keyword \kwd{CONST}, an identifier, an optional expression, and a
terminating semicolon:

$$\nonterm{const-decl} ::= \mbox{\kwd{CONST} \nonterm{identifier}
  \opt{\nonterm{expr}} \kwd{;}}$$

\bfhead{Data declarations} A data declaration consists of the keyword
\kwd{DATA}, an optional identifier representing a label for the data,
an expression or a string constant, and a terminating semicolon.  A
backslash may optionally follow the string constant, indicating that
the string is unterminated (see
Section~\ref{sec:code-gen:global-decls}).

$$\nonterm{data-decl} ::= \mbox{\kwd{DATA} \opt{\nonterm{identifier}}
  \group{\nonterm{expr} $|$ \group{\nonterm{string-const} \opt{\bs}}}
  \kwd{;}}$$

\bfhead{Variable declarations} A variable declaration consists of the
keyword \kwd{VAR}, an identifier representing the variable name, a
type, an optional initializer expression, and a terminating semicolon:

$$\nonterm{var-decl} ::= \mbox{\kwd{VAR} \nonterm{identifier} \kwd{:}
  \nonterm{type} \opt{\kwd{=} \nonterm{expr}} \kwd{;}}$$

\bfhead{Function declarations} A function declaration consists of the
keyword \kwd{FN}, an identifier representing the function name, the
function parameters enclosed in parentheses, an optional return type,
and the function body:

$$\nonterm{fn-decl} ::= \mbox{\kwd{FN} \nonterm{identifier} \kwd{(}
  \nonterm{fn-params} \kwd{)} \opt{\kwd{:} \nonterm{type}}
  \nonterm{fn-body}}$$

\noindent
The function parameters are a comma-separated list of zero or more
parameters:

$$\nonterm{fn-params} ::= \mbox{\opt{\nonterm{fn-param} \group{\kwd{,}
    \nonterm{fn-param}}$^*$}}$$

\noindent
A parameter consists of an identifier representing the parameter name
and type:

$$\nonterm{fn-param} ::= \mbox{\nonterm{identifier} \kwd{:}
  \nonterm{type}}$$

\noindent
A function body is either zero or more variable declarations and
statements enclosed in braces, or a semicolon indicating an externally
defined function:

$$\nonterm{fn-body} ::= \mbox{\kwd{\{} \nonterm{var-decl}$^*$
  \nonterm{stmt}$^*$ \kwd{\}} $|$ \kwd{;}}$$

\bfhead{Include declarations} An include declaration consists of the
keyword \kwd{INCLUDE}, a string constant, and a terminating semicolon:

$$\nonterm{include-decl} ::= \mbox{\kwd{INCLUDE}
  \nonterm{string-const} \kwd{;}}$$

\subsection{Statements}
\label{sec:syntax:statements}

A statement is an if statement, a while statement, a set statement, a
call statement, and increment statement, a decrement statement, return
statement, or a block statement:

$$\nonterm{stmt} ::= \mbox{\nonterm{if-stmt} $|$ \nonterm{while-stmt}
  $|$ \nonterm{set-stmt} $|$ \nonterm{call-stmt} $|$
  \nonterm{incr-stmt} $|$ \nonterm{decr-stmt} $|$
  \nonterm{return-stmt} $|$ \nonterm{block-stmt} }$$

\bfhead{If statements} An if statement consists of the keyword
\kwd{IF}, a conditional expression enclosed in parentheses, a
statement to execute if the condition is true, and optionally the
keyword \kwd{ELSE} followed by a statement to execute if the condition
is false:

$$\nonterm{if-stmt} ::= \mbox{\kwd{IF} \kwd{(} \nonterm{expr} \kwd{)}
  \nonterm{stmt} \opt{\kwd{ELSE} \nonterm{stmt}}}$$

\bfhead{While statements} A while statement consists of the keyword
\kwd{WHILE}, a test expression enclosed in parentheses, and a
statement to execute as long as the condition is true:

$$\nonterm{while-stmt} ::= \mbox{\kwd{WHILE} \kwd{(} \nonterm{expr}
  \kwd{)} \nonterm{stmt}}$$

\bfhead{Set statements} A set statement consists of the keyword
\kwd{SET}, an expression, an equals sign, a right-hand-side
expression, and a terminating semicolon:

$$\nonterm{set-stmt} ::= \mbox{\kwd{SET} \nonterm{expr} \kwd{=}
  \nonterm{expr} \kwd{;}}$$

\noindent
The left-hand side expression may not be a binary operation \kwd{=}.

\bfhead{Call statements} A call statement consists of a call
expression followed by a terminating semicolon:

$$\nonterm{call-stmt} ::= \mbox{\nonterm{call-expr} \kwd{;}}$$

\bfhead{Increment statements} An increment statement consists of the
keyword \kwd{INCR}, an expression, and a terminating semicolon:

$$\nonterm{incr-stmt} ::= \mbox{\kwd{INCR} \nonterm{expr} \kwd{;}}$$

\bfhead{Decrement statements} n decrement statement consists of the
keyword \kwd{DECR}, an expression, and a terminating semicolon:

$$\nonterm{decr-stmt} ::= \mbox{\kwd{DECR} \nonterm{expr} \kwd{;}}$$

\bfhead{Return statements} A return statement consists of the keyword
\kwd{RETURN}, an optional expression, and a terminating semicolon:

$$\nonterm{return-stmt} ::= \mbox{\kwd{RETURN} \opt{\nonterm{expr}}
    \kwd{;}}$$

\bfhead{Block statements} A block statement consists of zero or more
statements enclosed in braces:

$$\nonterm{block-stmt} ::= \mbox{\kwd{\{} \nonterm{stmt}$^*$
  \kwd{\}}}$$

\subsection{Expressions}
\label{sec:syntax:expressions}

An expression is an identifier, an indexed expression, a register
expression, a numeric constant, a call expression, a binary operation
expression, a unary operation expression, a parentheses expression, or
a typed expression.

\begin{equation*}
\begin{split}
%
\nonterm{expr} ::= & \mbox{\nonterm{identifier} $|$
  \nonterm{indexed-expr} $|$ \nonterm{register-expr} $|$
  \nonterm{numeric-const} $|$ \nonterm{call-expr} $|$} \\
%
  & \mbox{\nonterm{binop-expr} $|$ \nonterm{unop-expr} $|$
  \nonterm{parens-expr} $|$ \nonterm{typed-expr}}
%
\end{split}
\end{equation*}

\noindent
Except for binary expressions as specified below, any ambiguity in
parsing is resolved using right recursion.  For example, \kwd{@F()} is
parsed as \kwd{@(F())} and not \kwd{(@F)()}.

\bfhead{Numeric constants}  A numeric constant is an integer or
character constant:

$$\nonterm{numeric-const} ::= \mbox{\nonterm{integer-const} $|$
  \nonterm{char-const}}$$

\bfhead{Call expressions} A call expression consists of an expression
followed by an argument list enclosed in parentheses:

$$\nonterm{call-expr} ::= \mbox{\nonterm{expr} \kwd{(}
  \opt{\nonterm{expr} \group{\kwd{,} \nonterm{expr}}$^*$} \kwd{)}}$$

\bfhead{Indexed expressions} An indexed expression consists of a
base expression, an offset expression, and a type:

$$\nonterm{indexed-expr} ::= \mbox{\nonterm{expr} \kwd{[}
    \nonterm{expr} \kwd{,} \nonterm{type} \kwd{]}}$$

\bfhead{Register expressions} A register expression consists of a
caret character followed by a 6502 register name:

$$\nonterm{register-expr} ::= \mbox{\kwd{\textasciicircum} \group{\kwd{A} $|$
    \kwd{X} $|$ \kwd{Y} $|$ \kwd{P} $|$ \kwd{S}}}$$

\bfhead{Binary operation expressions} A binary operation expression
consists of a left-hand-side expression, a binary operator, and a
right-hand-side expression:

$$\nonterm{binop-expr} ::= \mbox{\nonterm{expr} \nonterm{binop} \nonterm{expr}}$$
%
$$\nonterm{binop} ::= \mbox{\kwd{>} $|$ \kwd{<} $|$ \kwd{<=} $|$
  \kwd{>=} $|$ \kwd{AND} $|$ \kwd{OR} $|$ \kwd{XOR} $|$ \kwd{+} $|$
  \kwd{-} $|$ \kwd{*} $|$ \kwd{/} $|$ \shl~$|$ \shr~$|$
  \kwd{=}}$$

\noindent
Parsing of binary operation expressions is disambiguated using the
following precedence rules:

\begin{enumerate}
%
\item In any sequence \nonterm{expr-1} \nonterm{binop-1}
  \nonterm{expr-2} \nonterm{binop-2} \nonterm{expr-3}, the implied
  parentheses go around the first operation unless the second
  operation has a strictly higher precedence.  There are five levels
  of precedence, from highest to lowest: (1) \kwd{<}\kwd{<} and \kwd{>}\kwd{>};
  (2) \kwd{*} and \kwd{\bs}; (3) \kwd{+} and \kwd{-}; (4) \kwd{=},
  \kwd{>}, \kwd{<}, \kwd{<=}, and \kwd{>=}; and (5) \kwd{AND},
  \kwd{OR}, and \kwd{XOR}.
%
\item Binary operators bind more weakly than any other expression
  combinators.  For example, \kwd{A+B()} is parsed as \kwd{A+(B())}
  and not \kwd{(A+B)()}; and \kwd{-5+3} is parsed as \kwd{(-5)+3} and
  not \kwd{-(5+3)}.
%
\end{enumerate}

\bfhead{Unary operation expressions} A unary operation expression
consists of a unary operator followed by an expression:

$$\nonterm{unop-expr} ::= \mbox{\nonterm{unop} \nonterm{expr}}$$

$$\nonterm{unop} ::= \mbox{\kwd{@} $|$ \kwd{NOT} $|$ \kwd{-}}$$

\bfhead{Parentheses expressions} A parentheses expression is an
expression surrounded by parentheses:

$$\nonterm{parens-expr} ::= \mbox{\kwd{(} \nonterm{expr} \kwd{)}}$$

\bfhead{Typed expressions} A typed expression is an expression with
an explicit type:

$$\nonterm{typed-expr} ::= \mbox{\nonterm{expr} \kwd{:}
  \nonterm{type}}$$

\subsection{Types}

A type is an unsigned size, a signed size, or an address type:

$$\nonterm{type} ::= \mbox{\nonterm{size} \opt{\kwd{S}} $|$ \kwd{@}}$$


\section{Semantic Checking}
\label{sec:semantics}

The compiler checks every source file for conformance to the semantic
rules stated below.  A semantically invalid program (i.e., one that
violates one of these rules) generates a compile-time error and causes
compilation to halt.  Only semantically valid files go on to code
generation as described in Section~\ref{sec:code-gen}.

\subsection{Representation of Values}
\label{sec:semantics:values}

The AppleCore language supports two types of values: integer values
and Boolean values.

\bfhead{Integer values} Integer values are represented in AppleCore
in the ordinary way for the Apple II:
%
\begin{enumerate}
%
\item Unsigned single-byte values are represented by treating the 8
  bits, low to high, as the 8 digits of a binary number.
%
\item Multi-byte values are stored in little-endian order.
%
\item Signed values are stored in two's complement representation
  (i.e., by constructing the absolute value of the number, flipping
  the bits, and adding one).
%
\end{enumerate}

\bfhead{Boolean values} The AppleCore language has no special Boolean
type; as in C, ordinary integers serve as Boolean values.  However,
AppleCore differs from C in what values it treats as ``true'' and
``false.''  In AppleCore, only the lowest-order bit is relevant in
testing the Boolean value of an integer: integers whose low bit is 1
(i.e., odd integers) all count as ``true,'' while integers whose low
bit is 0 (i.e., even integers) all count as ``false.''  This mechanism
allows bitwise \kwd{NOT} to function as logical \kwd{NOT}, without any
need for a separate logical negation operator, and that keeps the
language simple.  Also, the constants 1 and 0 can function as ``true''
and ``false.''  These names and values are not built into the
language, but one can easily declare constants \kwd{TRUE} and
\kwd{FALSE} with the values 1 and 0.

\subsection{Imported Declarations}
\label{sec:semantics:import}

Before performing any semantic checks, the compiler imports zero or
more declarations specified by the user, usually as a compiler option.
The option must allow the user to specify (1) the files from which to
read the declarations and (2) the order in which to read the files.

Each declaration so encountered is logically prepended to the source
file under translation, in the order encountered, subject to the
following rules:
%
\begin{enumerate}
%
\item Constant declarations are prepended as if they were
  literally included in the source file.
%
\item Function declarations are prepended as if they were literally
  included in the source file, but with a semicolon instead of a body
  (if any).
%
\item Variable and data declarations are prepended storing their name
  and size only; no variable initializer or data information is
  included for attribution (Section~\ref{sec:semantics:attr}), and
  these declarations generate no code (similar to an \kwd{extern}
  variable declaration in C).
%
\end{enumerate}

\subsection{Attribution}
\label{sec:semantics:attr}

The first semantic check performed by the compiler is to match each
identifier appearing in the source file with its corresponding
definition; this step is called \emph{attribution}.  Identifiers
appearing in imported variable and data declarations
(Section~\ref{sec:semantics:import}) do not appear in the source file
for purposes of this rule.  A \emph{definition} is a constant
declaration, data declaration, variable declaration, function
parameter declaration, or function declaration.  A \emph{use} is any
identifier appearing in the source file not as the identifier of a
definition.

The compiler matches definitions with uses in the following way:
%
\begin{enumerate}
%
\item At global scope (i.e., outside of any function body), for each
  use find the corresponding definition with the same identifier at
  global scope.  If there is no corresponding definition, or if any
  two definitions use the same identifier, then  report an error.
%
\item At function scope (i.e., inside a function body), for each use
  find the corresponding definition with the same identifier in the
  function scope, which consists of the global scope together with all
  function parameters and local variables declared in the function.
  If there is no corresponding definition, or if any two definitions
  in the function scope use the same identifier, then report an error.
%
\end{enumerate}
% 
Constant declarations must precede their uses in the source file.
Otherwise, declarations need not precede their uses.

\subsection{Constant-Value Expressions}
\label{sec:semantics:constant-value-expr}

The term ``constant value expression'' refers to either a compile-time
constant value expression or an assembly-time constant value
expression.

\bfhead{Compile-time constant value expressions} A compile-time
constant-value expression is one of the following:
%
\begin{enumerate}
%
\item A numeric constant.
%
\item An identifier corresponding to a constant declaration.
%
\item A binary operation expression, where each operand is a
  compile-time constant-value expression.
%
\item A negation (\kwd{-}) or \kwd{NOT} unary operation expression,
  where the operand is a compile-time constant-value expression.
%
\item A parentheses expression, where the expression inside the
  parentheses is a compile-time constant-value expression.
%
\item A typed expression, where the subexpression is a compile-time
  constant-value expression.
%
\end{enumerate}

\bfhead{Assembly-time constant value expressions} An assembly-time
constant value expression is one of the following:
%
\begin{enumerate}
%
\item A compile-time constant-value expression.
%
\item An identifier corresponding to a data declaration (useful for
  data tables).
%
\item An identifier corresponding to a function declaration (useful
  for jump tables).
%
\item An address expression \kwd{@}\nonterm{identifier}, where
  \nonterm{identifier} is an identifier corresponding to a global
  variable declaration (useful for creating a separate global symbol
  for the address of a global variable).
%
\end{enumerate}

\subsection{Expressions Appearing at Global Scope}
\label{sec:semantic:global-expr}

The following rules pertain to expressions appearing at global scope:
%
\begin{enumerate}
%
\item Any expression appearing in a constant declaration must be a
  compile-time constant-value expression
  (Section~\ref{sec:semantics:constant-value-expr}).
%
\item Any expression appearing in a data declaration or as the
  initializer of a global variable declaration must be an
  assembly-time constant-value expression
  (Section~\ref{sec:semantics:constant-value-expr}).
%
\end{enumerate}
%
These rules are designed so that the compiler can do a single pass
over all the expressions \nonterm{expr} appearing at global scope, in
the order in which they appear in the source file, and generate code
as described in Section~\ref{sec:code-gen:global-decls}.

\subsection{Size, Signedness, and Addresses}
\label{sec:semantics:size}

Every expression appearing in the source file is given a \emph{size}
and a \emph{signedness}.  An expression may (or may not)
\emph{represent an address}.  The address representation rules are
designed to exclude a class of easy-to-make yet potentially disastrous
errors such as dereferencing a variable's contents instead of its
address (for example, writing \kwd{X[0,1]} instead of
\kwd{(@X)[0,1]}).

\bfhead{Size and signedness associated with type} The type
\nonterm{size} with no trailing \kwd{S} has size \nonterm{size} and is
unsigned.  The type \nonterm{size} \kwd{S} has size \nonterm{size} and
is signed.  The type \kwd{@} has size 2 and is unsigned.

\bfhead{Computing the size and signedness of an expression} For a
constant-value typed expression
(Section~\ref{sec:semantics:constant-value-expr}), the size and
signedness come from the type.  For any other compile-time
constant-value expression, the size and signedness come from the value
produced by evaluating the expression as specified in
Section~\ref{sec:code-gen:expressions}, assuming that the operands to
any binary expression have the maximum allowed size of 255.  The size
is the minimum number of bytes required to represent the value.  The
expression is signed if the value is less than zero, otherwise
unsigned.

For any other expression, if the expression appears in a data
declaration or global variable declaration, then size is 2 and the
signedness is unsigned.  Otherwise the size and signedness are
computed as follows:
%
\begin{itemize}
%
\item \ithead{Identifiers} The size and signedness of an identifier
  come from its definition (Section~\ref{sec:semantics:attr}).  (1) If
  the definition is a variable declaration or function parameter
  declaration, then the size and signedness come from the type.  (2)
  If the definition is a data declaration or function declaration,
  then the size is 2, and the signedness is unsigned.
%
\item \ithead{Indexed expressions} The size and signedness of an
  indexed expression come from the \nonterm{type} component of the
  expression.
%
\item \ithead{Register expressions} The size of a register expression
  is 1.  It is unsigned.
%
\item \ithead{Call expressions} If the called expression (i.e., the
  expression before the first parenthesis) is anything other than an
  identifier corresponding to a function declaration, then the size is
  0 and the signedness is unsigned.  Otherwise the size and signedness
  come from the function declaration corresponding to the identifier:
  if the function has a return type, then the size and signedness come
  from the type; otherwise the size is 0 and the signedness is
  undefined.
%
\item \ithead{Binary operation expressions}  For comparison
  operations, the size is 1 byte unsigned.  For shift operations, the
  size and signedness come from the left operand.  For all other
  operations, the size is the maximum of the sizes of its operands.
  If either of the operands is signed, then the result is signed.
  Otherwise, the result is unsigned.
%
\item \ithead{Unary operation expressions} The size and signedness of
  a unary operation expression come from the operand, except that (1)
  an address operation \kwd{@} is two bytes unsigned; and (2) a
  negation operation is signed.
%
\item \ithead{Parentheses expressions} The size and signedness of a
  parentheses expression come from the enclosed expression.
%
\item \ithead{Typed expressions} The size and signedness of a typed
  expression come from its type.
%
\end{itemize}

\bfhead{Address expressions} The following expressions each represent
an address:
%
\begin{itemize}
%
\item \ithead{Identifiers} An identifier represents an address if its
  associated definition (1) is a constant declaration of size 1 or 2
  unsigned; or (2) is a data declaration; or (3) is a function
  declaration; or (4) is a variable or function parameter declaration
  whose type is \kwd{@}.
%
\item \ithead{Indexed expressions} An indexed expression represents an
  address if its \nonterm{type} component is \kwd{@}.
%
\item \ithead{Call expressions}  A call expression represents an
  address if the called expression is an identifier whose associated
  definition is a function with return type \kwd{@}.
%
\item \ithead{Binary expressions} A shift or divide operation
  represents an address if its left operand does.  A binary operation
  that is not a shift, divide, or compare represents an address if
  exactly one of its operands does.
%
\item \ithead{Unary operation expressions} An address operation
  \kwd{@} represents an address.  Otherwise, a unary operation
  represents an address if its operand does.
%
\item \ithead{Parentheses expressions} A parentheses expression
  represents an address if its subexpression does.
%
\item \ithead{Typed expressions} A typed expression represents an
  address if the type is \kwd{@}.
%
\end{itemize}
%
No other expression represents an address.

\bfhead{Size, signedness, and address requirements for expressions}
The following requirements apply to all expressions, whether or not
they are constant-value expressions:
%
\begin{itemize}
%
\item \ithead{Indexed expressions} The base expression (before the
  first bracket) of an indexed expression must represent an address.
  The index expression (immediately after the first bracket) must have
  size 1 or 2 and be unsigned.
%
\item \ithead{Call expressions} (1) The called expression must
  represent an address.  (2) Each argument expression must have
  nonzero size. (3) Each argument must satisfy the requirements for a
  valid address assignment (see below), treating the argument
  expression as the right-hand side of the assignment, and the
  corresponding function parameter as the left-hand side.
%
\item \ithead{Binary operation expressions}  For shift operations,
  the right-hand expression must be 1 byte unsigned.
%
\end{itemize}

\bfhead{Variable declarations} If an initializer expression is
present, then (1) the expression must have non-zero size; and (2) the
expression must satisfy the requirements for a a valid address
assignment, treating the variable declaration itself as the left-hand
side, and the initializer expression as the right-hand side.

\bfhead{Set statements} (1) The right-hand expression of a set
statement must have nonzero size.  (2) The assignment represented by
the statement must satisfy the requirements for a valid address
assignment.

\bfhead{Return statements} (1) No return statement may contain an
expression of zero size.  (2) If a function has nonzero return size,
and the function has a body, then (a) the body must end with a return
statement containing an expression, and (b) every return statement
appearing in the body must contain an expression.  (3) If a function
has return size 0, then no return statement in the function body may
contain an expression. (4) A return statement that contains an
expression must satisfy the requirements for a valid address
assignment, treating the contained expression as the right-hand side,
and treating the left-hand side as a fresh variable whose type is the
return type of the enclosing function.

\bfhead{Valid address assignments} (1) If the left-hand side of an
assignment represents an address, then the right-hand side must also
represent an address. (2) If the right-hand side of an assignment
represents and address and is not a constant-value expression, then
the left-hand side must also represent an address.

\bfhead{Function frame sizes} The \emph{frame size} of a function
declaration that contains a body equals the sum of the sizes of all
the function parameters and local variables.  No function may have a
frame size of greater than 255 bytes.

\subsection{Number of Function Arguments}
\label{sec:semantics:fn-args}

At every call expression where the called expression is an identifier
corresponding to a function declaration, the number of arguments must
match the number of parameters given in the declaration.

\subsection{Semantic LValues}
\label{sec:semantics:lvalues}

\bfhead{Definition of semantic lvalue}  A \emph{semantic lvalue} is
an expression that (1) is an identifier, register expression, or
indexed expression; and (2) if it is an identifier, then it
corresponds to a variable or function parameter declaration.

\bfhead{Requirement of semantic lvalues}  (1) A semantic lvalue is
required in the following places: (a) on the left-hand side of a set
statement; (b) as the expression of an increment or decrement
statement; and (c) as the operand of an address (\kwd{@}) unary
operator.  If an expression that is not a semantic lvalue appears in
any of these places, then a compile-time error results.  (2) A
compile-time error results if a register expression appears as the
operand of an address (\kwd{@}) operator.


\section{Code Generation}
\label{sec:code-gen}

\subsection{Source Files}
\label{sec:code-gen:source-files}

The AppleCore language is designed to support separate compilation.
To do that, the compiler translates source files to assembly files in
one of two modes: \emph{top-level mode} and \emph{include mode}.  The
mode must be specified at the time the source file is translated to
assembly, usually as a compiler option.

\bfhead{Top-level mode}  In top-level mode, the compiler translates
the source file as follows:
%
\begin{enumerate}
%
\item Issue some code (or an assembler directive to include the code)
  that sets up the program stack
  (Section~\ref{sec:code-gen:program-stack}).
%
\item Issue code to transfer control to the first function with a body
  that appears in the file.  There must be at least one such function,
  and it is a compile-time error if not.
%
\item Translate all global declarations in the source file as
  described in Section~\ref{sec:code-gen:global-decls}, in the
  order in which they appear in the source file.  Include declarations
  may appear anywhere in the file, and they are translated to
  directives to include the corresponding assembly files.
%
\item Issue assembler directives to include the AppleCore support code
  (for example, the code needed to do arithmetic operations).
%
\end{enumerate}
%

\bfhead{Include mode} In include mode, the compiler just translates
the declarations in the source file as described in
Section~\ref{sec:code-gen:global-decls}.

\bfhead{Final assembly and symbol resolution} A user program consists
of exactly one file translated in top-level mode and zero or more
files translated in include mode.  The final step in compilation is to
ask the assembler to assemble the file translated in top-level mode,
which includes all the other files for assembly.  This assembly also
causes name resolution for externally defined names.  It is the user's
responsibility to ensure that externally-defined names are both
available and uniquely defined; the compiler cannot check this.  If a
symbol used in one of the assembled files is not defined in any of the
files (or if any symbol is multiply defined), then an assembler error
will result.

\subsection{Global Declarations}
\label{sec:code-gen:global-decls}

Global declarations are translated as stated below, except for
imported declarations (Section~\ref{sec:semantics:import}), which
generate no code.  To the extent that translation produces actual
bytes of code, data, or reserved storage, the bytes must be laid out
in the order in which the translated constructs are encountered in the
source file.

\bfhead{Constant declarations} If the constant declaration
\nonterm{const-decl} contains an expression \nonterm{expr}, then the
compiler evaluates \nonterm{expr} to a numeric constant according to
the rules in Section~\ref{sec:code-gen:expressions}.  In the rest of
the source file where \nonterm{const-decl} appears, the compiler
replaces each instance of the identifier given in \nonterm{const-decl}
expression with the value so computed.  The compiler does \emph{not}
generate a label or other assembly artifact associated with
\nonterm{const-decl}.  In particular, the same identifier may appear
in different constant declarations in different files that are part of
the same final assembly (Section~\ref{sec:code-gen:source-files}).

If the constant declaration contains no expression, then the name in
the declaration must refer to the two-byte unsigned value associated
with an assembler label during final assembly
(Section~\ref{sec:code-gen:source-files}).

\bfhead{Data declarations} (1) If an expression appears before the
semicolon, then the compiler evaluates the expression to an assembler
expression according to the rules in
Section~\ref{sec:code-gen:expressions} and generates code to store the
bytes of the constant in little-endian order.  (2) If a string
constant appears before the semicolon, then the compiler generates
assembly code to store the bytes of the string constant in
left-to-right order.  (a) If no backslash \kwd{\bs} follows the string
constant, then the compiler adds a terminating NUL (\$00) character
after the last byte of string data. (b) If a backslash follows the
string constant, then no NUL character is added.  (3) In any case, if
an identifier appears after the keyword \kwd{DATA}, then the compiler
associates the name with the lowest address in which data is stored.

\bfhead{Global variable declarations} The compiler sets aside a number
of bytes of storage equal to the declared size of the variable and
associates the identifier appearing after \kwd{VAR} with the lowest
address of this storage.  If an expression appears before the
semicolon, then the compiler evaluates the expression to an assembler
expression according to the rules stated in
Section~\ref{sec:code-gen:expressions} and issues assembly code to
fill in the storage set aside for the variable with the bytes of the
numeric constant, in little-endian order, using size adjustment (if
necessary) as stated in Section~\ref{sec:code-gen:size-adjustment}.

\bfhead{Function declarations} (1) A function declaration with no
body does not correspond to any generated code; its signature is used
only for checking call expressions that invoke the function.  (2) A
function declaration with a body causes code to be generated as stated
in Section~\ref{sec:code-gen:fn-bodies}.

\bfhead{Include declarations}  The compiler translates an include
declaration into an assembler directive to include the file named in
the string constant.

\subsection{Program Stack}
\label{sec:code-gen:program-stack}

This specification requires that the implementation provide three
logical stacks for use at runtime, collectively referred to as the
\emph{program stack}:
%
\begin{enumerate}
%
\item A C-style \emph{call stack}, such that a new frame appears at
  the top of the stack for each function call, and the frame is popped
  at the end of the call.  This stack is used in generating function
  call code, as specified in Section~\ref{sec:code-gen:fn-bodies}.
%
\item An \emph{expression stack} for use during expression evaluation
  as specified in Section~\ref{sec:code-gen:expressions}.
%
\item An \emph{allocation stack} that supports an \kwd{ALLOCATE}
  library function that works as follows.  (1) \kwd{ALLOCATE} is only
  ever called when no temporary expression result is on the expression
  stack; otherwise the results are undefined.  (2) When \kwd{ALLOCATE}
  is called, it is passed a two-byte size value $s$.  (3) (a) If there
  is enough space on the stack, then \kwd{ALLOCATE} returns a pointer
  to $s$ bytes on the stack. (b) If there are not $s$ bytes remaining
  on the stack, then the results are undefined. (4) The $s$ bytes
  pointed to by the result of the \kwd{ALLOCATE} call remain valid
  (i.e., are not modified except by user-written code) during the
  lifetime of the call stack frame for the function in which the
  \kwd{ALLOCATE} call occurred.  (5) At the end of the function in
  which the \kwd{ALLOCATE} call occurred, the memory is deallocated
  and may be reused as program stack storage.
%
\end{enumerate}

The compiler may provide this stack functionality in any way that is
feasible.  Typically the same physical stack would be used for all
three logical stacks (and the logical stacks have been specified so
that this is possible), but this is not a requirement.  Also,
typically the generated code would maintain its own two-byte stack
pointer, because the native Apple II stack located at addresses \$200
through \$2FF is too small to serve as the program stack for this
language.  However, this specification does not mandate any particular
stack size; the actual available stack size depends on both the
compiler implementation and the program memory requirements (for
example, whether the graphics and/or language card areas are available
for program heap and stack data).

\subsection{Function Bodies}
\label{sec:code-gen:fn-bodies}

The AppleCore language definition assumes a C-style function call
stack as specified in Section~\ref{sec:code-gen:program-stack}.  For
each function declaration that has a body, the compiler does the
following:
%
\begin{itemize}
%
\item Generate code to store the callee's frame pointer on the call
  stack.
%
\item Generate code to reserve slots on the call stack for the
  function parameters, local variables, and saved registers.  The
  saved registers are all the registers appearing in register
  expressions in the function body.
%
\item For each local variable declaration that includes an initializer
  expression, generate code to evaluate the expression according to
  the rules in Section~\ref{sec:code-gen:expressions} and assign the
  resulting value into the stack slot for the variable, using size
  adjustment (if necessary) as stated in
  Section~\ref{sec:code-gen:size-adjustment}.
%
\item For each statement of the function body, generate code as stated
  in Section~\ref{sec:code-gen:statements}.  The statements of a
  function body are executed in sequence, in the order in which they
  appear in the program text.
%
\item If the last statement executed in the function is not a return
  statement, then generate code to pop the frame off the stack and
  restore the parent frame.
%
\end{itemize}
%
This calling convention is designed so that from the point of view of
the caller, a call to an AppleCore function that takes no arguments
and returns no result is indistinguishable from a \kwd{JSR} to a
non-AppleCore function (for example, in the System Monitor).  That
way, non-AppleCore assembly language functions can be called in the
same way as AppleCore functions.  In particular, if \kwd{V} is a
variable, then a call expression \kwd{V()} has the same meaning
regardless of what kind of function address is stored in \kwd{V}
(which is not generally known at compile time).  Also, a source file
with a function declaration \kwd{FN FOO();} can be assembled with any
assembly file containing a function with label \kwd{FOO}, regardless
of whether the code at label \kwd{FOO} uses the AppleCore conventions
for managing the stack.

\subsection{Statements}
\label{sec:code-gen:statements}

The compiler generates code to do the following for each program
statement.

\bfhead{If statements} Evaluate the conditional expression
(Section~\ref{sec:code-gen:expressions}), pop the result off the
expression stack, and mask off all but the lowest bit of the result to
get the corresponding Boolean value
(Section~\ref{sec:semantics:values}).  If the Boolean value of the
result is 1 (true), then execute the true part.  If the Boolean value
of the result is 0 (false), then either do nothing (if there is no
optional \kwd{ELSE} clause) or execute the statement in the \kwd{ELSE}
clause (if one is provided).

\bfhead{While statements} Evaluate the conditional expression
(Section~\ref{sec:code-gen:expressions}), pop the result off the
expression stack, and mask off all but the lowest bit of the result to
get the corresponding Boolean value
(Section~\ref{sec:semantics:values}).  If the Boolean value of the
result is 1 (true), then execute the body of the loop and repeat the
whole process.  If the Boolean value of the result is 0 (false), then
do nothing.

\bfhead{Set statements}  (1) Evaluate the right-hand side expression,
adjusting the size if necessary
(Section~\ref{sec:code-gen:size-adjustment}) to conform to the size of
the left-hand side expression.  (2) The left-hand side expression must
be a semantic lvalue (Section~\ref{sec:semantics:lvalues}). Compute
its address as specified in Section~\ref{sec:code-gen:lvalues}.  (3)
Copy the result of step (1) into memory starting at the address
computed in step (2).

\bfhead{Call statements} Evaluate the expression contained in the
statement (Section~\ref{sec:code-gen:expressions}).  If it produces a
result of nonzero size, pop the result off the stack.

\bfhead{Increment statements} The operand expression must be a
semantic lvalue. (Section~\ref{sec:semantics:lvalues}).  Compute its
address as stated in Section~\ref{sec:code-gen:lvalues}.  Add one to
the value stored at the address and store the result in place.

\bfhead{Decrement statements} The operand expression must be a
semantic lvalue. (Section~\ref{sec:semantics:lvalues}).  Compute its
address as stated in Section~\ref{sec:code-gen:lvalues}.  Subtract one
from the value stored at the address and store the result in place.

\bfhead{Return statements} (1) If the statement contains an
expression, then evaluate the expression
(Section~\ref{sec:code-gen:expressions}) and ensure that it appears at
the top of the expression stack
(Section~\ref{sec:code-gen:program-stack}) on function return.  (2)
Pop the current frame off the call stack and restore the parent frame.

\bfhead{Block statements} Execute each statement of the block in the
order in which it appears in the program text.

\subsection{Expressions}
\label{sec:code-gen:expressions}

We define the semantics of expressions using an expression stack, as
specified in Section~\ref{sec:code-gen:program-stack}.  The compiler
generates code to do the following for each program expression.

\bfhead{Identifiers} (1) If the identifier is a semantic lvalue
(Section~\ref{sec:semantics:lvalues}), then compute its address as
specified in Section~\ref{sec:code-gen:lvalues} and get a number of
bytes equal to the size of the right-hand side
(Section~\ref{sec:semantics:size}), starting at that address.  Push
those bytes on the stack, using size adjustment if necessary
(Section~\ref{sec:code-gen:size-adjustment}) to generate a result that
has a number of bytes equal to the left-hand expression size. (2) If
the identifier corresponds to a numeric constant declaration, then
push the bytes of the value associated with the declaration
(Section~\ref{sec:code-gen:global-decls}) on the stack in
little-endian order.  (3) If the identifier corresponds to a data
declaration, then push the two-byte address associated with the
identifier (Section~\ref{sec:code-gen:global-decls}) on the stack in
little-endian order. (4) If the identifier corresponds to a function
declaration, then push the two-byte address corresponding to the first
byte of the function body (Section~\ref{sec:code-gen:fn-bodies}) on
the stack in little-endian order.  (If the function body is defined in
a different file, this can be done by using the label associated with
the function body; at assembly time, the label definition will be
available.)

\bfhead{Numeric constants} Push the bytes of the value associated
with the declaration (Section~\ref{sec:code-gen:global-decls}) on the
stack in little-endian order.

\bfhead{Indexed expressions} (1) Compute the address of the indexed
expression as stated in Section~\ref{sec:code-gen:lvalues}.  (2) Get a
number of bytes equal to the size specified in the expression,
starting at that address, and push them on the stack.

\bfhead{Register expressions} Push the one-byte value stored in the
function call stack slot for that register
(Section~\ref{sec:code-gen:fn-bodies}) on the expression stack.

\bfhead{Call expressions} The procedure depends on the kind of the
called expression.

\ithead{(1) Declared functions:} If the called expression is an
identifier corresponding to a function declaration, then do the
following.  (a) Evaluate each argument and store it to its slot on the
stack in the new frame, adjusting the argument sizes
(Section~\ref{sec:code-gen:size-adjustment}) as necessary to them to
the parameter sizes specified in the function definition. (b) Adjust
the stack pointer to point to the base of the new frame.  (c) Restore
all the saved registers (Section~\ref{sec:code-gen:fn-bodies}) to the
values saved in their slots on the stack.  (d) Do a \kwd{JSR} to the
code for the function body (Section~\ref{sec:code-gen:fn-bodies}). (e)
Save all the saved registers to their slots.

\ithead{(2) Compile-time constants:} If the called expression
evaluates to a compile-time constant, then (a) restore the registers
as in (1)(c), (b) \kwd{JSR} to the constant address, and (c) save the
registers as in (1)(e).

\ithead{(3) Callee address unknown:} Otherwise the callee address is
not known at compile time.  In this case proceed as in case 2, but in
step (2)(b) store the value to zero-page memory and do a \kwd{JSR} to
an indirect \kwd{JMP} to the address.

The rules are designed so that in all cases the last value written
into the stack slot for each saved register prior to the call is
guaranteed to be in the corresponding 6502 register just before the
\kwd{JSR} to the called function.  Thus, except for the \kwd{S}
register, all saved registers are guaranteed to have their saved
values at the point where control enters the callee function.  Because
of the \kwd{JSR}, the \kwd{S} register's value will be increased by 2.
Similarly, all stack slots corresponding to saved registers other than
the \kwd{S} register will contain the values that the corresponding
registers had on exit from the function; the \kwd{S} register's value
will be off by 2.

\bfhead{Binary operation expressions} 
%
\ithead{Shift operations (\shl~and \shr)} Evaluate the left-hand side,
evaluate the right-hand side, pop the right-hand side, and shift the
left-hand side value (which is the result on the top of the stack) by
a number of bytes equal to the right-hand side value.  In the case of
a right shift, the shift is signed (arithmetic) if the left-hand side
expression is signed; otherwise it is unsigned (logical).

\ithead{Arithmetic operations (\kwd{*}, \kwd{/}, \kwd{+}, and
  \kwd{-})} Evaluate the left and right expressions, adjust their
size if necessary (Section~\ref{sec:code-gen:size-adjustment}) to make
their sizes equal to the size of the entire expression, pop the
results, perform the operation, and push the result.  In the case of
multiplication and division, the operation is signed if the entire
expression is signed, otherwise unsigned.  Because AppleCore uses
two's complement representation for negative numbers
(Section~\ref{sec:semantics:values}), there is no distinction between
signed and unsigned addition and subtraction.  The division operation
produces the integer quotient.  All operations are done without regard
to overflow (i.e., if overflow occurs, they produce the low-order
bits, up to the size of the expression).

\ithead{Comparison operations (\kwd{=}, \kwd{>}, \kwd{<}, \kwd{>=},
  and \kwd{<=})} Evaluate the left and right expressions, adjust
their size if necessary (Section~\ref{sec:code-gen:size-adjustment})
to make their sizes equal to the size of the entire expression, pop
the results, perform the operation, and push the result.  The result
is a single-byte value such that 1 represents true and 0 represents
false.  In the case of comparisons other than equality, the operation
is signed if the entire expression is signed, otherwise unsigned.

\ithead{Bitwise operations (\kwd{AND}, \kwd{OR}, and \kwd{XOR})}
Evaluate the left and right expressions, adjust their size if
necessary (Section~\ref{sec:code-gen:size-adjustment}) to make their
sizes equal to the size of the entire expression, pop the results,
perform the operation, and push the result.

\bfhead{Unary operation expressions} 
%
\ithead{Dereference (\kwd{@})} The operand expression must be a
semantic lvalue (Section~\ref{sec:semantics:lvalues}).  Compute and
push its address as stated in Section~\ref{sec:code-gen:lvalues}.

\ithead{Arithmetic negation (\kwd{-}) and bitwise not (\kwd{NOT})}
Evaluate the operand expression and perform the operation on it.
Negation is two's complement negation (i.e., \kwd{NOT} plus 1).

\bfhead{Parentheses expressions} Evaluate the expression inside the
parentheses.

\bfhead{Typed expressions} Evaluate the subexpression.  Adjust the
size to generate a new value of the correct size for the given type.

\subsection{Address Computation for Semantic LValues}
\label{sec:code-gen:lvalues}

Computation of the address of a semantic lvalue \nonterm{expr} works
as follows:
%
\begin{enumerate}
%
\item If \nonterm{expr} is an identifier corresponding to a global
  variable, then its address is the lowest address of the global
  storage assigned to that variable
  (Section~\ref{sec:code-gen:global-decls}). 
%
\item If \nonterm{expr} is an identifier corresponding to a function
  parameter or local variable, then its address is the lowest address
  of the function call stack slot assigned to that parameter or
  variable (Section~\ref{sec:code-gen:fn-bodies}).
%
\item If \nonterm{expr} is an indexed expression, then its address is
  computed by evaluating the base expression, evaluating the offset
  expression, popping and adding the results, and adjusting the size
  of the result to 2 if necessary
  (Section~\ref{sec:code-gen:size-adjustment}). 
%
\item If \nonterm{expr} is a register expression, then its address is
  the address of the function call stack slot assigned to that
  register (Section~\ref{sec:code-gen:fn-bodies}).
%
\end{enumerate}


\subsection{Size Adjustment}
\label{sec:code-gen:size-adjustment}

\emph{Size adjustment} applies whenever a result is assigned into or
out of a semantic lvalue and the sizes of the source and target values
do not match.  The following rules apply to size adjustment:
%
\begin{enumerate}
%
\item If the source value size is greater than the target value size,
  then high-order bytes of the source value are \emph{truncated}: only
  the low-order bytes of the source value are stored, up to the number
  of bytes required by the target value.
%
\item If the source value size is smaller than the target value size,
  and the source value is unsigned, then the source value is
  \emph{zero extended}:  the bytes of the source value form the
  low-order bytes of the target value, and zeros are used to fill the
  rest of the bytes of the target value.
%
\item If the source value size is smaller than the target value size,
  and the source value is signed, then the source value is \emph{sign
    extended}: the bytes of the source value form the low-order bytes
  of the target value, and the rest of the bytes of the target value
  are filed with either all 0 bits (if the high-order bit of the
  source value is 0) or all 1 bits (if the high-order bit of the
  source value is 1).
%
\end{enumerate}


\end{document}
